---
title: "curate tax assignments and decon ASVs"
output: html_document
date: "2025-12-18"
author: "Kimberly Ledger"
---

```{r}
library(tidyverse)
library(ggplot2)
library(here)
```

perform additional taxonomic curation and summary of ASV taxonomic assignments
```{r}
## load taxonomic assignments
assignments <- read.csv(here("data", "20260520_mifish_tax_outputs", "collapsed_taxonomy.csv")) %>% unique() %>% select(!X)

## load LULU output
map <- read.csv(here("data", "20260520_mifish_lulu_ref", "asvmap_lulu98.csv")) %>%
  rename(ASV = X)

## merge 
join <- map %>%
  left_join(assignments, by = "ASV") %>%
  rename(ASV_taxon = taxon) %>%
  select(ASV:curated, ASV_taxon) %>%
  left_join(assignments, by = c("parent_id" = "ASV")) %>%
  rename(parent_taxon = taxon) %>%
  select(ASV:ASV_taxon, parent_taxon)

mismatches <- join %>%
  filter(ASV_taxon != parent_taxon)

mismatches_species <- mismatches %>%
  filter(str_detect(ASV_taxon, " "))
# keep species-level tax assignments that lulu merged to higher classification...
# in other words, use ASV_taxon here

keeper_1 <- mismatches_species$ASV
## otherwise, use the parent ASV id  
```

update asv to use for taxonomic assignment 
```{r}
map_updated <- map %>%
  select(!spread) %>%
  select(!rank) %>%
  mutate(parent_id = ifelse(ASV %in% keeper_1, ASV, parent_id),
         curated = ifelse(ASV %in% keeper_1, "updated", curated))
```

merge taxonomic assignments with ASV table 
```{r}
asv_table <- read.csv(here("data", "20260520_mifish_outputs", "ASVtable.csv")) %>%
  rename(sample_id = X) %>%
  pivot_longer(cols = -1, names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(map, by = "ASV") %>%                  ### change this if using lulu map or map_updated
  #left_join(map_updated, by = "ASV") %>%
  select(!total) %>%
  select(!curated) %>%
  rename(ASV_for_tax_ID = parent_id) %>%
  left_join(assignments, by = c("ASV_for_tax_ID" = "ASV"))
```


### start decontamination steps ### 

#### 1 - remove ASVs that do not have a taxonomic assingment 
```{r}
asv_table_filter1 <- asv_table %>%
  filter(!is.na(taxon))

removed_asvs <- anti_join(asv_table, asv_table_filter1) %>%
  group_by(ASV) %>%
  summarize(total_reads = sum(reads)) %>%
  arrange(desc(total_reads))
removed_asvs

#how many ASVs and total reads removed because of no tax assignment? 
nrow(removed_asvs)
sum(removed_asvs$total_reads)
#proportion of reads removed because of no taxonomic assignment
sum(removed_asvs$total_reads)/sum(asv_table$reads)
```

look at non-fish taxonomic assignments 
```{r}
asv_table_filter1 %>% filter(class != "Actinopteri") %>% filter(reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = sum(reads),
            n_pcr_reps = n_distinct(sample_id))
```

remove terrestrial non-fish taxonomic assignments
```{r}
asv_table_filter2 <- asv_table_filter1 %>%
  filter(!taxon %in% c("Bos", "Bos taurus", "Bovidae", "Canis lupus", "Felis catus", "Homo sapiens", "Ovis", "Rangifer tarandus", "Sus scrofa")) 
```


add column to sample tracking table to include number of reads per sample after taxonomic assignment 
```{r}
track <- read.csv(here("data", "20260520_mifish_outputs", "track.csv")) %>%
  rename(sample_id = X) %>%
  rename(fastq_ID = sample_id) %>% 
  mutate(fastq_ID = gsub("-", "_", fastq_ID))

asv_table_filter2 <- asv_table_filter2 %>%
  rename(fastq_ID = sample_id) %>% 
  mutate(fastq_ID = gsub("-", "_", fastq_ID))

reads_per_sample <- asv_table_filter2 %>%
  group_by(fastq_ID) %>%
  summarize(tax_reads = sum(reads))

track <- track %>%
  left_join(reads_per_sample) %>%
  mutate(tax_reads = ifelse(is.na(tax_reads), 0, tax_reads))
```


load sample metadata 
```{r}
metadata <- read.csv(here("data", "metadata_20260520run.csv")) %>%
  filter(primer == "MiFish")
```

add column to the ASV table that labels the sample type
```{r}
asv_table_with_sample_type <- metadata %>%
  dplyr::select(fastq_ID, extraction_ID, replicate, protocol) %>%
  left_join(asv_table_filter2, by = "fastq_ID")
```

#### 2 - account for tag-jumping using the positive controls 

first, plot
```{r}
asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  ggplot(aes(x=interaction(extraction_ID, replicate), y=reads, fill=ASV_for_tax_ID)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  ggplot(aes(x=interaction(extraction_ID, replicate), y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

identify the maximum proportion of reads for each ASV found in the positive controls
```{r}
prop_asvs_in_positives <- asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  group_by(extraction_ID, replicate) %>%
  mutate(TotalReadsPerSample = sum(reads)) %>%
  mutate(Prop = reads/TotalReadsPerSample) %>%
  group_by(ASV_for_tax_ID) %>%
  summarise(max_prop = max(Prop)) %>%
  arrange(desc(max_prop)) 
prop_asvs_in_positives
```

*** very little tag-jumping so won't filter for now. 


### 3 - Account for contaminants in negative controls 

plot negatives 

first, pcr negatives 
```{r}
asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "NC") 
```

nothing in PCR negative controls! 
no other negatives used in this test set 

20260520 - very little in PCR negative


#### 4 - Remove low read depth samples based on ASV accumulation curve

```{r}
library(vegan)

asv_table_wide <- asv_table_with_sample_type %>%
  select(extraction_ID, replicate, protocol, ASV_for_tax_ID, reads) %>%
  filter(!is.na(reads)) %>%
  mutate(reads = as.integer(reads)) %>%
  group_by(extraction_ID, replicate, protocol, ASV_for_tax_ID) %>%
  summarise(reads = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = ASV_for_tax_ID, values_from = reads, values_fill = 0)

extraction_ID <- asv_table_wide$extraction_ID

asv_table_wide <- asv_table_wide %>%
  ungroup() %>%
  select(-extraction_ID) %>%
  select(-replicate) %>%
  select(-protocol)

## plots the figure
rarecurve(asv_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of ASVs Identified",
          xlim = c(0,3000))
```


summarize in a table how many pcr replicates meet certain read count thresholds 
```{r}
read_summary <- asv_table_with_sample_type %>%
  filter(!is.na(reads)) %>%
  group_by(extraction_ID, replicate, protocol) %>%
  summarize(tot_reads = sum(reads)) %>%
  arrange(desc(tot_reads)) %>%
  group_by(protocol) %>%
  summarize(atleast0 = sum(tot_reads >= 0),
            atleast100 = sum(tot_reads >= 100),
            atleast500 = sum(tot_reads >= 500),
            atleast1k = sum(tot_reads >= 1000),
            atleast2k = sum(tot_reads >= 2000))
read_summary
```


apply a threshold of 1000 reads per pcr replicate
```{r}
asv_table_filter3 <- asv_table_with_sample_type %>%
  group_by(extraction_ID, replicate, protocol) %>%
  mutate(tot_reads = sum(reads)) %>%
  filter(tot_reads >= 1000)
```

```{r}
saveRDS(asv_table_filter3, file = here("data", "20260520_mifish_decon", "decontaminated_asv_table.RDS"))
```

also update and output the sample tracking file 
```{r}
reads_per_sample_decon <- asv_table_filter3 %>%
  group_by(fastq_ID) %>%
  summarize(decon_reads = sum(reads)) 

track <- track %>%
  left_join(reads_per_sample_decon, by = "fastq_ID") %>%
  mutate(decon_reads = ifelse(is.na(decon_reads), 0, decon_reads))

write.csv(track, here("data", "20260520_mifish_decon", "track.csv"))
```
