---
title: "MiFish summary"
output: html_document
date: "2026-04-01"
---

libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
#devtools::install_github("thomasp85/scico")
library(scico)
library(vegan)
library(reshape2)
```

load decontaminated ASV table 
```{r}
asv_table <- readRDS(here("data", "mifish_decon", "decontaminated_asv_table.RDS"))
```

convert to taxon table
make a few additional taxon edits to not have nested assignments 
and remove any remaining pc reads 
```{r}
taxon_table <- asv_table %>%
  filter(taxon != "Acipenser fulvescens") %>% 
  mutate(taxon = ifelse(taxon == "Atheresthes evermanni", "Atheresthes", taxon),
         taxonomic_level = ifelse(taxon == "Atheresthes", "genus", taxonomic_level),
         species = ifelse(taxon == "Atheresthes", NA, species),
         taxon = ifelse(taxon == "Hemitripterus", "Hemitripterus bolini", taxon),
         taxonomic_level = ifelse(taxon == "Hemitripterus bolini", "species", taxonomic_level),
         species = ifelse(taxon == "Hemitripterus bolini", "Hemitripterus bolini", species)) %>%
  #mutate(taxon = ifelse(taxon == "Clupea", "Clupea pallasii", taxon),
  #       species = ifelse(taxon == "Clupea pallasii", "Clupea pallasii", species),
  #       taxonomic_level = ifelse(taxon == "Clupea pallasii", "species", taxonomic_level),
  #       taxon = ifelse(taxon == "Hydrolagus", "Hydrolagus colliei", taxon),
  #       species = ifelse(taxon == "Hydrolagus colliei", "Hydrolagus colliei", species),
  #       taxonomic_level = ifelse(taxon == "Hydrolagus colliei", "species", taxonomic_level),
  #       taxon = ifelse(taxon == "Leuroglossus", "Leuroglossus schmidti", taxon),
  #       species = ifelse(taxon == "Leuroglossus schmidti", "Leuroglossus schmidti", species),
  #       taxonomic_level = ifelse(taxon == "Leuroglossus schmidti", "species", taxonomic_level),
  #      taxon = ifelse(taxon == "Microstomus", "Microstomus pacificus", taxon),
  #       species = ifelse(taxon == "Microstomus pacificus", "Microstomus pacificus", species),
  #      taxonomic_level = ifelse(taxon == "Microstomus pacificus", "species", taxonomic_level),
  #       taxon = ifelse(taxon == "Oncorhynchus gorbuscha", "Oncorhynchus", taxon),
  #       taxon = ifelse(taxon == "Oncorhynchus kisutch", "Oncorhynchus", taxon),
  #       species = ifelse(taxon == "Oncorhynchus", NA, species),
  #       taxonomic_level = ifelse(taxon == "Oncorhynchus", "genus", taxonomic_level),
  #       taxon = ifelse(taxon == "Sebastes nigrocinctus", "Sebastes", taxon),
  #       species = ifelse(taxon == "Sebastes", NA, species),
   #      taxonomic_level = ifelse(taxon == "Sebastes", "genus", taxonomic_level),
   #      taxon = ifelse(taxon == "Stenobrachius", "Stenobrachius leucopsarus", taxon),
   #      species = ifelse(taxon == "Stenobrachius leucopsarus", "Stenobrachius leucopsarus", species),
   #      taxonomic_level = ifelse(taxon == "Stenobrachius leucopsarus", "species", taxonomic_level)) %>%
  group_by(extraction_ID, replicate, protocol, taxon, taxonomic_level, species, genus, family, order, class) %>%
  summarize(reads = sum(reads)) %>%
  filter(reads > 0) %>%
  filter(extraction_ID != "PC")
```

load sample metadata
```{r}
metadata <- read.csv(here("data", "metadata_20260325run.csv")) %>%
#metadata <- read.csv(here("data", "metadata_20260520run.csv")) %>%
  select(extraction_ID, station) %>%
  unique()
# 
# # get depth from catch data 
# catch <- read.csv(here("catch_outputs", "metaprobe_haul_catch.csv")) %>%
#   select(haul, depth_m) %>% unique()
# 
# taxon_table_w_meta <- taxon_table %>% 
#   left_join(metadata) %>%
#   left_join(catch, by = "haul")
```

summary table to compare new vs old longmire extraction protocol
```{r}
protocol_summary <- taxon_table %>% 
  left_join(metadata, by = "extraction_ID") %>% 
  group_by(protocol, extraction_ID, replicate, station) %>%
  summarize(reads_per_replicate = sum(reads)) %>%
  group_by(protocol) %>%
  summarise(n_tech = n_distinct(extraction_ID, replicate),
            n_extract = n_distinct(extraction_ID),
            n_station = n_distinct(station),
            avg_reads_per_replicate = mean(reads_per_replicate),
            sd_reads_per_replicate = sd(reads_per_replicate))
```

make a summary table of the species and reads
```{r}
taxon_table_summary <- taxon_table %>%
  left_join(metadata, by = "extraction_ID") %>%
  group_by(protocol, taxon, species, genus, family, order) %>%
  summarize(total_reads = sum(reads),
            n_tech = n_distinct(extraction_ID, replicate),
            n_extract = n_distinct(extraction_ID),
            n_stations = n_distinct(station)) %>%
  arrange(taxon)

taxon_table_summary %>% group_by(protocol) %>% summarise(n_taxa = n_distinct(taxon))

taxa_compare <- taxon_table_summary %>%
  select(protocol, taxon, n_tech, n_extract, n_stations) %>%
  pivot_wider(
    names_from = protocol,
    values_from = c(n_tech, n_extract, n_stations),
    values_fill = 0
  )

#taxa_compare <- taxa_compare %>%
#  mutate(
#    diff_stations = n_stations_new - n_stations_old,
#    diff_extracts = n_extract_new - n_extract_old,
#    diff_tech = n_tech_new - n_tech_old
#  ) %>% ungroup()

#taxa_compare %>%
#  summarise(
#    mean_diff_stations = mean(diff_stations),
#    mean_diff_extracts = mean(diff_extracts),
#    mean_diff_tech = mean(diff_tech)
#  )

ggplot(taxa_compare, aes(x = n_stations_old, y = n_stations_new)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Old protocol", y = "New protocol")
```

```{r}
taxa_long <- taxa_compare %>%
  pivot_longer(
    cols = c(n_tech_old, n_tech_new,
             n_extract_old, n_extract_new,
             n_stations_old, n_stations_new),
    names_to = c("level", "protocol"),
    names_pattern = "n_(.*)_(old|new)",
    values_to = "n"
  )

taxa_plot <- taxa_long %>%
  mutate(level = factor(level,
                        levels = c("stations", "extract", "tech"),
                        labels = c("Station",  "Extraction", "Technical"))) %>%
  pivot_wider(names_from = protocol, values_from = n) %>%
  group_by(level, old, new) %>%
  summarise(n_taxa = n(), .groups = "drop") %>%
  ggplot(aes(x = old, y = new, color = level, size = n_taxa)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  theme_bw() +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  scale_size_continuous(range = c(1, 8), breaks = c(1, 3, 7))
taxa_plot
```
```{r}
#ggsave(here("figures", "mifish_taxon_detections.png"), taxa_plot, dpi = 300, height = 4, width = 6)
```


```{r}
taxon_table %>%
  left_join(metadata, by = "extraction_ID") %>%
  group_by(protocol, station) %>%
  summarize(n_taxa = n_distinct(taxon)) %>%
  ungroup() %>%
  group_by(protocol) %>%
  summarize(min = min(n_taxa),
            mean = mean(n_taxa),
            max = max(n_taxa))
```


## summarize the technical replicates by extraction

fill in 0's for taxa and then calculate means 
*** looking at this code later, i;m not sure this doing what i intended it to.... 
```{r}
taxon_table_mean <- taxon_table %>%
  ungroup() %>%
  select(extraction_ID, replicate, protocol, taxon, reads) %>%
  pivot_wider(names_from = taxon, values_from = reads, values_fill = 0) %>%
  pivot_longer(cols = c(4:57), names_to = "taxon", values_to = "reads") %>%       ### update based on number of species
#  pivot_longer(cols = c(4:20), names_to = "taxon", values_to = "reads") %>%
  group_by(extraction_ID, replicate, protocol) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ungroup() %>%
  group_by(extraction_ID, replicate, protocol, taxon) %>%
  summarize(mean_prop = mean(prop_reads))

#check each sums to 1 
taxon_table_mean %>% group_by(extraction_ID, replicate, protocol) %>% summarise(total = sum(mean_prop))
```

### eastern GOA

plot eastern GOA stations 
```{r}
station_meta <- read.csv(here("data", "2022_GOA_eDNA_stations.csv"))

eGOA <- station_meta %>% filter(StartLongitude > -140)

scico_pal <- scico(15, palette = 'roma')

eGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% eGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  ) + 
  guides(fill = guide_legend(ncol = 2))
eGOA_stations 

#ggsave(here("figures", "20260520_eGOA_stations.png"), plot =eGOA_stations, dpi = 300, width = 9, height = 5)
```

```{r}
eGOA_taxon_summary <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% eGOA$Station) %>%
  # remove taxa with zero total reads across the filtered data
  group_by(taxon) %>%
  filter(sum(mean_prop) > 0) %>%
  ungroup() %>%
  group_by(taxon) %>%
  summarise(
    # number of extractions with detection
    n_extract = n_distinct(extraction_ID[mean_prop > 0]),
    # number of stations with detection
    n_stations = n_distinct(station[mean_prop > 0]),
    # mean read proportion per station
    mean_prop_per_station = mean(tapply(mean_prop, station, sum)),
    .groups = "drop"
  ) %>%
  arrange(desc(n_stations)) %>%
  mutate(taxon = factor(taxon, levels = taxon))

eGOA_taxa_plot <- ggplot(eGOA_taxon_summary, aes(x = taxon, y = n_stations)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_bw() +
  labs(
    x = "Taxon",
    y = "Number of Stations with Detection"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
eGOA_taxa_plot

#ggsave(here("figures", "eGOA_taxa.png"), plot = eGOA_taxa_plot, dpi = 300, width = 9, height = 5)
```

## central GOA
```{r}
cGOA <- station_meta %>% filter(StartLongitude < -140) %>% filter(StartLongitude > -151)

scico_pal <- scico(10, palette = 'roma')

cGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% cGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
cGOA_stations 

ggsave(here("figures", "20260520_cGOA_stations.png"), plot =cGOA_stations, dpi = 300, width = 9, height = 5)
```

```{r}
cGOA_taxon_summary <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% cGOA$Station) %>%
  # remove taxa with zero total reads across the filtered data
  group_by(taxon) %>%
  filter(sum(mean_prop) > 0) %>%
  ungroup() %>%
  group_by(taxon) %>%
  summarise(
    # number of extractions with detection
    n_extract = n_distinct(extraction_ID[mean_prop > 0]),
    # number of stations with detection
    n_stations = n_distinct(station[mean_prop > 0]),
    # mean read proportion per station
    mean_prop_per_station = mean(tapply(mean_prop, station, sum)),
    .groups = "drop"
  ) %>%
  arrange(desc(n_stations)) %>%
  mutate(taxon = factor(taxon, levels = taxon))

cGOA_taxa_plot <- ggplot(cGOA_taxon_summary, aes(x = taxon, y = n_stations)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_bw() +
  labs(
    x = "Taxon",
    y = "Number of Stations with Detection"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
cGOA_taxa_plot

#ggsave(here("figures", "cGOA_taxa.png"), plot = cGOA_taxa_plot, dpi = 300, width = 9, height = 5)
```

#WGOA
```{r}
wGOA <- station_meta %>% filter(StartLongitude < -151)

scico_pal <- scico(5, palette = 'roma')

wGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% wGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
wGOA_stations 

ggsave(here("figures", "20260520_wGOA_stations.png"), plot =wGOA_stations, dpi = 300, width = 9, height = 3)
```


```{r}
allGOA_taxon_summary <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  # remove taxa with zero total reads across the filtered data
  group_by(taxon) %>%
  filter(sum(mean_prop) > 0) %>%
  ungroup() %>%
  group_by(taxon) %>%
  summarise(
    # number of extractions with detection
    n_extract = n_distinct(extraction_ID[mean_prop > 0]),
    # number of stations with detection
    n_stations = n_distinct(station[mean_prop > 0]),
    # mean read proportion per station
    mean_prop_per_station = mean(tapply(mean_prop, station, sum)),
    .groups = "drop"
  ) %>%
  arrange(desc(n_stations)) %>%
  mutate(taxon = factor(taxon, levels = taxon))

allGOA_taxa_plot <- ggplot(allGOA_taxon_summary, aes(x = taxon, y = n_stations)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_bw() +
  labs(
    x = "Taxon",
    y = "Number of Stations with Detection"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
allGOA_taxa_plot

ggsave(here("figures", "20260520_allGOA_taxa.png"), plot = allGOA_taxa_plot, dpi = 300, width = 9, height = 5)
```


### using just the old extraction protocol samples, generate species accumulation curves to estimate number of PCR replicates needed to asympote

```{r}
library(vegan)

old_taxon_table <- taxon_table_mean %>%
  left_join(metadata) %>%
  ungroup() %>%
  filter(protocol == "old") %>%
  mutate(pcr_rep = paste(extraction_ID, replicate, sep = "_")) %>% 
  select(station, extraction_ID, pcr_rep, taxon, mean_prop) %>%
  pivot_wider(names_from = "taxon", values_from = "mean_prop")

taxa_cols <- setdiff(
  names(old_taxon_table),
  c("station", "extraction_ID", "pcr_rep")
)

cum_richness <- old_taxon_table %>%
  #group_by(extraction_ID) %>%
  group_by(station) %>%
  arrange(pcr_rep, .by_group = TRUE) %>%
  group_modify(~{

    nrep <- nrow(.x)

    data.frame(
      replicates = seq_len(nrep),
      richness = sapply(seq_len(nrep), function(i){

        subset <- .x[1:i, taxa_cols]

        sum(colSums(subset > 0) > 0)

      })
    )

  })
```

```{r}
summary_curve <- cum_richness %>%
  group_by(replicates) %>%
  summarise(
    mean_richness = mean(richness),
    sd_richness = sd(richness),
    n = n()
  )

ggplot(summary_curve,
       aes(replicates, mean_richness)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_richness - sd_richness,
      ymax = mean_richness + sd_richness
    ),
    width = 0.1
  ) +
  labs(
    x = "Number of PCR replicates",
    y = "Mean cumulative richness"
  ) +
  theme_bw()
```


gain from each PCR rep
```{r}
# metadata columns
meta_cols <- c("station", "extraction_ID", "pcr_rep")

# taxa columns
taxa_cols <- setdiff(names(old_taxon_table), meta_cols)

cum_richness <- old_taxon_table %>%
  arrange(station,extraction_ID, pcr_rep) %>%
  #group_by(extraction_ID) %>%
  group_by(station) %>%
  group_modify(~{

    nrep <- nrow(.x)

    tibble(
      n_reps = seq_len(nrep),
      richness = map_dbl(seq_len(nrep), function(i){

        subset <- .x[1:i, taxa_cols]

        sum(colSums(subset > 0) > 0)

      })
    )

  }) %>%
  ungroup()

richness_gains <- cum_richness %>%
  #group_by(extraction_ID) %>
  group_by(station) %>%
  arrange(n_reps) %>%
  mutate(
    gain = richness - lag(richness),
    pct_gain = 100 * gain / lag(richness)
  ) %>%
  ungroup()

## absolute gains 
richness_gains %>%
  filter(n_reps > 1) %>%
  group_by(n_reps) %>%
  summarise(
    mean_gain = mean(gain, na.rm = TRUE),
    sd_gain = sd(gain, na.rm = TRUE),
    n = n()
  )

## percent of final gains 
final_richness <- cum_richness %>%
  #group_by(extraction_ID) %>%
  group_by(station) %>%
  summarise(
    final_richness = max(richness)
  )

prop_detected <- cum_richness %>%
#  left_join(final_richness, by = "extraction_ID") %>%
    left_join(final_richness, by = "station") %>%
  mutate(
    pct_final = 100 * richness / final_richness
  )

prop_detected %>%
  group_by(n_reps) %>%
  summarise(
    mean_pct = mean(pct_final),
    sd_pct = sd(pct_final)
  )
```

### another appraoch - How much of the community detected with all 3 PCR replicates would have been recovered if I had only run 1 or 2 PCR replicates?


```{r}
taxa_cols <- setdiff(
  names(old_taxon_table),
  c("station", "extraction_ID", "pcr_rep")
)

pa_table <- old_taxon_table %>%
  mutate(
    across(all_of(taxa_cols), ~ as.integer(. > 0))
  )

# function to pool PCR replicates
pool_reps <- function(df, n_reps, taxa_cols){

  chosen <- sample(seq_len(nrow(df)), n_reps)

  pooled <- colSums(df[chosen, taxa_cols]) > 0

  as.numeric(pooled)
}
```

calculate similarity to full dataset
```{r}
library(vegan)
library(dplyr)
library(purrr)

calc_similarity <- function(df, n_reps, taxa_cols){

  # full community (all PCRs)
  full_comm <- as.numeric(
    colSums(df[, taxa_cols]) > 0
  )

  # random subset of PCRs
  subset_comm <- pool_reps(df, n_reps, taxa_cols)

  # Jaccard similarity
  sim <- 1 - vegdist(
    rbind(full_comm, subset_comm),
    method = "jaccard",
    binary = TRUE
  )

  as.numeric(sim)
}
```

bootstrap
```{r}
set.seed(123)

n_iter <- 1000

results <- map_dfr(
  seq_len(n_iter),
  function(i){

    pa_table %>%
      group_by(extraction_ID) %>%
      group_modify(~{

        n_available <- nrow(.x)

        tibble(
          pcrs = c(1,2),
          similarity = c(
            calc_similarity(.x,
                            n_reps = 1,
                            taxa_cols = taxa_cols),

            if(n_available >= 2)
              calc_similarity(.x,
                              n_reps = 2,
                              taxa_cols = taxa_cols)
            else
              NA_real_
          )
        )

      }) %>%
      ungroup()

  }
)
```

summarize
```{r}
summary_results <- results %>%
  group_by(pcrs) %>%
  summarise(
    mean_similarity = mean(similarity,
                           na.rm = TRUE),
    sd_similarity = sd(similarity,
                       na.rm = TRUE),
    lower95 = quantile(similarity,
                       0.025,
                       na.rm = TRUE),
    upper95 = quantile(similarity,
                       0.975,
                       na.rm = TRUE)
  )

summary_results
```


could also do this with:  vegdist(..., method = "bray") on pooled mean_prop values instead of pres/abs 

```{r}
pool_reps_bc <- function(df, n_reps, taxa_cols){

  chosen <- sample(seq_len(nrow(df)), n_reps)

  colMeans(df[chosen, taxa_cols], na.rm = TRUE)
}

calc_bray_similarity <- function(df, n_reps, taxa_cols){

  # full community (all PCR replicates averaged)
  full_comm <- colMeans(df[, taxa_cols], na.rm = TRUE)

  # subset community (1 or 2 PCR replicates averaged)
  subset_comm <- pool_reps_bc(df, n_reps, taxa_cols)

  # Bray-Curtis distance → convert to similarity
  sim <- 1 - vegdist(
    rbind(full_comm, subset_comm),
    method = "bray"
  )

  as.numeric(sim)
}
```


```{r}
set.seed(123)

n_iter <- 1000

results_bray <- map_dfr(
  seq_len(n_iter),
  function(i){

    old_taxon_table %>%
      group_by(extraction_ID) %>%
      group_modify(~{

        n_available <- nrow(.x)

        tibble(
          pcrs = c(1, 2),
          similarity = c(
            calc_bray_similarity(.x, 1, taxa_cols),

            if(n_available >= 2)
              calc_bray_similarity(.x, 2, taxa_cols)
            else
              NA_real_
          )
        )

      }) %>%
      ungroup()

  }
)
```

summarize
```{r}
summary_bray <- results_bray %>%
  group_by(pcrs) %>%
  summarise(
    mean_similarity = mean(similarity, na.rm = TRUE),
    sd = sd(similarity, na.rm = TRUE),
    q025 = quantile(similarity, 0.025, na.rm = TRUE),
    q975 = quantile(similarity, 0.975, na.rm = TRUE)
  )

summary_bray
```

```{r}
get_full_station <- function(df, taxa_cols){

  colMeans(df[, taxa_cols], na.rm = TRUE)

}

sample_station_pcrs <- function(df, n, taxa_cols){

  chosen <- sample(seq_len(nrow(df)), n)

  colMeans(df[chosen, taxa_cols], na.rm = TRUE)
}

calc_station_bray <- function(df, n_pcr, taxa_cols){

  full_comm <- get_full_station(df, taxa_cols)

  subset_comm <- sample_station_pcrs(df, n_pcr, taxa_cols)

  1 - as.numeric(
    vegdist(
      rbind(full_comm, subset_comm),
      method = "bray"
    )
  )
}

set.seed(123)

n_iter <- 1000

results_station <- map_dfr(
  seq_len(n_iter),
  function(i){

    old_taxon_table %>%
      group_by(station) %>%
      group_modify(~{

        n_available <- nrow(.x)

        tibble(
          pcrs = c(1, 2, 3, 4, 5),
          similarity = map_dbl(1:5, function(n){

            if(n_available >= n){
              calc_station_bray(.x, n, taxa_cols)
            } else {
              NA_real_
            }

          })
        )

      }) %>%
      ungroup()

  }
)
```

```{r}
summary_station <- results_station %>%
  group_by(pcrs) %>%
  summarise(
    mean_similarity = mean(similarity, na.rm = TRUE),
    sd = sd(similarity, na.rm = TRUE),
    q025 = quantile(similarity, 0.025, na.rm = TRUE),
    q975 = quantile(similarity, 0.975, na.rm = TRUE),
    n = n()
  )

summary_station
```

