---
title: "curate tax assignments and decon ASVs"
output: html_document
date: "2026-04-01"
author: "Kimberly Ledger"
---

```{r}
library(tidyverse)
library(ggplot2)
library(here)
```

perform additional taxonomic curation and summary of ASV taxonomic assignments
```{r}
## load taxonomic assignments
assignments <- read.csv(here("data", "rkfish_outputs", "taxonomy_customrkfishdb.csv")) %>% unique() %>% select(!X) %>%
  rename(ASV = qseqid)
```

any advantage to using lulu with the rockfish seqs??? skipping for now. 

merge taxonomic assignments with ASV table 
```{r}
asv_table <- read.csv(here("data", "rkfish_outputs", "ASVtable.csv")) %>%
  rename(sample_id = X) %>%
  pivot_longer(cols = -1, names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(assignments, by = "ASV")
```


### start decontamination steps ### 

#### 1 - remove ASVs that do not have a taxonomic assingment 
```{r}
asv_table_filter1 <- asv_table %>%
  filter(!is.na(taxon))

removed_asvs <- anti_join(asv_table, asv_table_filter1) %>%
  group_by(ASV) %>%
  summarize(total_reads = sum(reads)) %>%
  arrange(desc(total_reads))
removed_asvs

#how many ASVs and total reads removed because of no tax assignment? 
nrow(removed_asvs)
sum(removed_asvs$total_reads)
#proportion of reads removed because of no taxonomic assignment
sum(removed_asvs$total_reads)/sum(asv_table$reads)
```


add column to sample tracking table to include number of reads per sample after taxonomic assignment 
```{r}
track <- read.csv(here("data", "rkfish_outputs", "track.csv")) %>%
  rename(sample_id = X) #%>%
  #rename(fastq_ID = sample_id) %>% 
  #mutate(sample_id = gsub("-", "_", sample_id))

reads_per_sample <- asv_table_filter1 %>%
  group_by(sample_id) %>%
  summarize(tax_reads = sum(reads))

track <- track %>%
  left_join(reads_per_sample) %>%
  mutate(tax_reads = ifelse(is.na(tax_reads), 0, tax_reads))
```


load sample metadata 
```{r}
metadata <- read.csv(here("data", "metadata_20260325run.csv")) %>%
  filter(primer == "Rockfish")
```

add column to the ASV table that labels the sample type
```{r}
asv_table_filter1 <- asv_table_filter1 %>%
  rename(fastq_ID = sample_id) %>% 
  mutate(fastq_ID = gsub("-", "_", fastq_ID))

asv_table_with_sample_type <- metadata %>%
  dplyr::select(fastq_ID, extraction_ID, replicate, protocol) %>%
  left_join(asv_table_filter1, by = "fastq_ID")
```

#### 2 - account for tag-jumping using the positive controls 

first, plot - PC was S.carnatus (gopher rockfish)
```{r}
asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  ggplot(aes(x=interaction(extraction_ID, replicate), y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  ggplot(aes(x=interaction(extraction_ID, replicate), y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

identify the maximum proportion of reads for each ASV found in the positive controls
```{r}
prop_asvs_in_positives <- asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "PC") %>%
  group_by(extraction_ID, replicate) %>%
  mutate(TotalReadsPerSample = sum(reads)) %>%
  mutate(Prop = reads/TotalReadsPerSample) %>%
  group_by(ASV) %>%
  summarise(max_prop = max(Prop)) %>%
  arrange(desc(max_prop)) 
prop_asvs_in_positives
```

subtract the max proportion of tag-jumped reads for each ASV from samples
```{r}
indexhop_table <- asv_table_with_sample_type %>%
  group_by(extraction_ID, replicate) %>%
  mutate(TotalReadsPerSample = sum(reads, na.rm = T)) %>%
  left_join(prop_asvs_in_positives, by = c("ASV")) %>%
  mutate(IndexHoppingReads = TotalReadsPerSample*max_prop) %>%
  mutate(IndexHoppingReads = if_else(is.na(IndexHoppingReads), 0, IndexHoppingReads)) %>%
  mutate(IndexHoppingReads = as.integer(IndexHoppingReads)) %>%
  mutate(reads_IndexHop_removed = reads - IndexHoppingReads) %>%
  mutate(reads_IndexHop_removed = if_else(reads_IndexHop_removed < 0, 0, reads_IndexHop_removed))
head(indexhop_table)
```

looks like these asvs don't actually show up in the environmental sample very much, but will include the filter for good measure anyways..
clean up the table by removing columns no longer needed 
```{r}
asv_table_filter2 <- indexhop_table %>%
  dplyr::select(!reads) %>%
  dplyr::select(!TotalReadsPerSample) %>%
  dplyr::select(!max_prop) %>%
  dplyr::select(!IndexHoppingReads) %>%
  dplyr::rename(reads = reads_IndexHop_removed)
```


### 3 - Account for contaminants in negative controls 

plot negatives 

first, pcr negatives 
```{r}
asv_table_filter2  %>%
  filter(reads > 0) %>%
  filter(extraction_ID == "NC") 
```

lot of rosethorn (Sebastes helvomaculatus) in one PCR negative control replicate


#### 4 - Remove low read depth samples based on ASV accumulation curve

```{r}
library(vegan)

asv_table_wide <- asv_table_filter2 %>%
  select(extraction_ID, replicate, protocol, ASV, reads) %>%
  filter(!is.na(reads)) %>%
  mutate(reads = as.integer(reads)) %>%
  group_by(extraction_ID, replicate, protocol, ASV) %>%
  summarise(reads = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = ASV, values_from = reads, values_fill = 0)

extraction_ID <- asv_table_wide$extraction_ID

asv_table_wide <- asv_table_wide %>%
  ungroup() %>%
  select(-extraction_ID) %>%
  select(-replicate) %>%
  select(-protocol)

## plots the figure
rarecurve(asv_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of ASVs Identified",
          xlim = c(0,3000))
```


summarize in a table how many pcr replicates meet certain read count thresholds 
```{r}
read_summary <- asv_table_filter2 %>%
  filter(!is.na(reads)) %>%
  group_by(extraction_ID, replicate, protocol) %>%
  summarize(tot_reads = sum(reads)) %>%
  arrange(desc(tot_reads)) %>%
  group_by(protocol) %>%
  summarize(atleast0 = sum(tot_reads >= 0),
            atleast100 = sum(tot_reads >= 100),
            atleast500 = sum(tot_reads >= 500),
            atleast1k = sum(tot_reads >= 1000),
            atleast2k = sum(tot_reads >= 2000))
read_summary
```


apply a threshold of 1000 reads per pcr replicate
```{r}
asv_table_filter3 <- asv_table_with_sample_type %>%
  group_by(extraction_ID, replicate, protocol) %>%
  mutate(tot_reads = sum(reads)) %>%
  filter(tot_reads >= 1000)
```

```{r}
#saveRDS(asv_table_filter3, file = here("data", "rkfish_decon", "decontaminated_asv_table.RDS"))
```

also update and output the sample tracking file 
```{r}
reads_per_sample_decon <- asv_table_filter3 %>%
  group_by(fastq_ID) %>%
  summarize(decon_reads = sum(reads)) 

track <- track %>%
  rename(fastq_ID = sample_id) %>% 
  mutate(fastq_ID = gsub("-", "_", fastq_ID)) %>%
  left_join(reads_per_sample_decon, by = "fastq_ID") %>%
  mutate(decon_reads = ifelse(is.na(decon_reads), 0, decon_reads))

#write.csv(track, here("data", "rkfish_decon", "track.csv"))
```
