---
title: "Rockfish summary"
output: html_document
date: "2026-04-01"
---

libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
#devtools::install_github("thomasp85/scico")
library(scico)
library(vegan)
library(reshape2)
```

load decontaminated ASV table 
```{r}
asv_table <- readRDS(here("data", "rkfish_decon", "decontaminated_asv_table.RDS"))
```

convert to taxon table
make a few additional taxon edits to not have nested assignments 
and remove any remaining pc reads 
```{r}
taxon_table <- asv_table %>%
  filter(taxon != "Sebastes carnatus") %>% 
  group_by(extraction_ID, replicate, protocol, taxon) %>%
  summarize(reads = sum(reads)) %>%
  filter(reads > 0) %>%
  filter(extraction_ID != "PC") %>%
  filter(extraction_ID != "NC")
```

load sample metadata
```{r}
metadata <- read.csv(here("data", "metadata_20260325run.csv")) %>%
  select(extraction_ID, station) %>%
  unique()
# 
# # get depth from catch data 
# catch <- read.csv(here("catch_outputs", "metaprobe_haul_catch.csv")) %>%
#   select(haul, depth_m) %>% unique()
# 
# taxon_table_w_meta <- taxon_table %>% 
#   left_join(metadata) %>%
#   left_join(catch, by = "haul")
```

summary table to compare new vs old longmire extraction protocol
```{r}
protocol_summary <- taxon_table %>% 
  left_join(metadata, by = "extraction_ID") %>% 
  group_by(protocol, extraction_ID, replicate, station) %>%
  summarize(reads_per_replicate = sum(reads)) %>%
  group_by(protocol) %>%
  summarise(n_tech = n_distinct(extraction_ID, replicate),
            n_extract = n_distinct(extraction_ID),
            n_station = n_distinct(station),
            avg_reads_per_replicate = mean(reads_per_replicate),
            sd_reads_per_replicate = sd(reads_per_replicate))
```

make a summary table of the species and reads
```{r}
taxon_table_summary <- taxon_table %>%
  left_join(metadata, by = "extraction_ID") %>%
  group_by(protocol, taxon) %>%
  summarize(total_reads = sum(reads),
            n_tech = n_distinct(extraction_ID, replicate),
            n_extract = n_distinct(extraction_ID),
            n_stations = n_distinct(station)) %>%
  arrange(taxon)

taxon_table_summary %>% group_by(protocol) %>% summarise(n_taxa = n_distinct(taxon))

taxa_compare <- taxon_table_summary %>%
  select(protocol, taxon, n_tech, n_extract, n_stations) %>%
  pivot_wider(
    names_from = protocol,
    values_from = c(n_tech, n_extract, n_stations),
    values_fill = 0
  )

#taxa_compare <- taxa_compare %>%
#  mutate(
#    diff_stations = n_stations_new - n_stations_old,
#    diff_extracts = n_extract_new - n_extract_old,
#    diff_tech = n_tech_new - n_tech_old
#  ) %>% ungroup()

#taxa_compare %>%
#  summarise(
#    mean_diff_stations = mean(diff_stations),
#    mean_diff_extracts = mean(diff_extracts),
#    mean_diff_tech = mean(diff_tech)
#  )

ggplot(taxa_compare, aes(x = n_stations_old, y = n_stations_new)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Old protocol", y = "New protocol")
```

```{r}
taxa_long <- taxa_compare %>%
  pivot_longer(
    cols = c(n_tech_old, n_tech_new,
             n_extract_old, n_extract_new,
             n_stations_old, n_stations_new),
    names_to = c("level", "protocol"),
    names_pattern = "n_(.*)_(old|new)",
    values_to = "n"
  )

taxa_plot <- taxa_long %>%
  mutate(level = factor(level,
                        levels = c("stations", "extract", "tech"),
                        labels = c("Station",  "Extraction", "Technical"))) %>%
  pivot_wider(names_from = protocol, values_from = n) %>%
  group_by(level, old, new) %>%
  summarise(n_taxa = n(), .groups = "drop") %>%
  ggplot(aes(x = old, y = new, color = level, size = n_taxa)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  theme_bw() +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  scale_size_continuous(range = c(1, 5), breaks = c(1, 2, 3))
taxa_plot
```
```{r}
#ggsave(here("figures", "rkfish_taxon_detections.png"), taxa_plot, dpi = 300, height = 4, width = 6)
```


```{r}
taxon_table %>%
  left_join(metadata, by = "extraction_ID") %>%
  group_by(protocol, station) %>%
  summarize(n_taxa = n_distinct(taxon)) %>%
  ungroup() %>%
  group_by(protocol) %>%
  summarize(min = min(n_taxa),
            mean = mean(n_taxa),
            max = max(n_taxa))
```

## summarize the technical replicates by extraction

fill in 0's for taxa and then calculate means 
```{r}
taxon_table_mean <- taxon_table %>%
  ungroup() %>%
  select(extraction_ID, replicate, protocol, taxon, reads) %>%
  pivot_wider(names_from = taxon, values_from = reads, values_fill = 0) %>%
  pivot_longer(cols = c(4:22), names_to = "taxon", values_to = "reads") %>%
  group_by(extraction_ID, replicate, protocol) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ungroup() %>%
  group_by(extraction_ID, replicate, protocol, taxon) %>%
  summarize(mean_prop = mean(prop_reads))

#check each sums to 1 
taxon_table_mean %>% group_by(extraction_ID, replicate, protocol) %>% summarise(total = sum(mean_prop))
```

### eastern GOA

plot eastern GOA stations 
```{r}
station_meta <- read.csv(here("data", "2022_GOA_eDNA_stations.csv"))

eGOA <- station_meta %>% filter(StartLongitude > -140)

scico_pal <- scico(19, palette = 'roma')

eGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% eGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
eGOA_stations 

ggsave(here("figures", "eGOA_stations_rockfish.png"), plot =eGOA_stations, dpi = 300, width = 9, height = 5)
```


## central GOA
```{r}
cGOA <- station_meta %>% filter(StartLongitude < -140) %>% filter(StartLongitude > -151)

scico_pal <- scico(8, palette = 'roma')

cGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% cGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
cGOA_stations 

#ggsave(here("figures", "cGOA_stations_rockfish.png"), plot =cGOA_stations, dpi = 300, width = 9, height = 5)
```

#WGOA
```{r}
wGOA <- station_meta %>% filter(StartLongitude < -151)

scico_pal <- scico(3, palette = 'roma')

wGOA_stations <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  filter(station %in% wGOA$Station) %>%
  group_by(taxon) %>%
  mutate(
    pres = ifelse(sum(mean_prop) > 0, 1, 0),
    # create a simplified x-axis label
    protocol_label = ifelse(protocol == "old", "O", "N"),
    # still keep unique x for each extractionID/replicate/protocol combo
    x_id = interaction(extraction_ID, replicate, protocol, drop = TRUE)
  ) %>%
  filter(pres == 1) %>%
  ggplot(aes(x=interaction(extraction_ID, replicate, protocol), y = mean_prop, fill = taxon)) +
  geom_bar(stat = "identity") + 
  # replace axis labels with protocol_label
  scale_x_discrete(labels = function(x) {
    # x comes in as the levels of x_id; map them to protocol_label
    protocol_map <- setNames(ifelse(grepl("old", x), "O", "N"), x)
    protocol_map[x]
  }) +
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~station, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
wGOA_stations 

#ggsave(here("figures", "wGOA_stations_rockfish.png"), plot =wGOA_stations, dpi = 300, width = 9, height = 3)
```


```{r}
allGOA_taxon_summary <- taxon_table_mean %>%
  left_join(metadata, by = "extraction_ID") %>%
  # remove taxa with zero total reads across the filtered data
  group_by(taxon) %>%
  filter(sum(mean_prop) > 0) %>%
  ungroup() %>%
  group_by(taxon) %>%
  summarise(
    # number of extractions with detection
    n_extract = n_distinct(extraction_ID[mean_prop > 0]),
    # number of stations with detection
    n_stations = n_distinct(station[mean_prop > 0]),
    # mean read proportion per station
    mean_prop_per_station = mean(tapply(mean_prop, station, sum)),
    .groups = "drop"
  ) %>%
  arrange(desc(n_stations)) %>%
  mutate(taxon = factor(taxon, levels = taxon))

allGOA_taxa_plot <- ggplot(allGOA_taxon_summary, aes(x = taxon, y = n_stations)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_bw() +
  labs(
    x = "Taxon",
    y = "Number of Stations with Detection"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
allGOA_taxa_plot

#ggsave(here("figures", "allGOA_taxa_rockfish.png"), plot = allGOA_taxa_plot, dpi = 300, width = 9, height = 5)
```

