---
title: "station map"
output: html_document
date: "2026-04-01"
---

make a map of GOA 2022 Coral Stations

libraries 
```{r}
library(tidyverse)
library(akgfmaps)
library(sf)
library(ggplot2)
library(here)
```


```{r}
#browseVignettes('akgfmaps')
#?make_idw_map
?get_base_layers

goa <- get_base_layers(select.region = "goa", set.crs = 3338)
```

```{r}
eDNA_test <- read.csv(here("data", "metadata_20260325run.csv"))

stations <- read.csv(here("data", "2022_GOA_eDNA_stations.csv")) %>%
  select(Station, StartLatitude, StartLongitude, MaximumDepthInMeters, Temperature, Substrate) %>%
  unique() %>%
  mutate(MaximumDepthInMeters = as.numeric(MaximumDepthInMeters)) %>%
  mutate(test = ifelse(Station %in% eDNA_test$station, "yes", "no"))

# Step 1: Convert to sf object (default CRS is WGS84, EPSG:4326)
stations_sf <- st_as_sf(stations, coords = c("StartLongitude", "StartLatitude"), crs = 4326)

# Step 2: Transform to Alaska Albers Equal Area (EPSG:3338)
stations_sf_3338 <- st_transform(stations_sf, crs = 3338)

# Print the resulting spatial object
print(stations_sf_3338)
```

```{r}
goa_y <- c(510000, 1200000)
goa_x <- c(-350000, 1400000)  

goa_plot <- ggplot() +
  #geom_sf(data = goa$bathymetry) +
  #geom_sf(data = goa$survey.strata) + 
  geom_sf(data = goa$akland, fill = "grey70") +
  geom_sf(data = stations_sf_3338, aes(color = MaximumDepthInMeters), size = 1) +
  scale_color_viridis_c(direction = -1, name = "Maximum Depth (m)") +
  coord_sf(xlim = goa_x,
           ylim = goa_y) +
  theme_bw()
goa_plot
#ggsave(here("figures", "station_map.png"), plot = goa_plot, dpi = 300, width = 8, height = 3)
```

```{r}
goa_test_plot <- ggplot() +
  geom_sf(data = goa$bathymetry) +
  geom_sf(data = goa$akland, fill = "grey70") +
  geom_sf(data = stations_sf_3338, aes(color = test), size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  coord_sf(xlim = goa_x,
           ylim = goa_y) +
  theme_bw() + 
  labs(color = "In Test?")
goa_test_plot
#ggsave(here("figures", "test_station_map.png"), plot = goa_test_plot, dpi = 300, width = 8, height = 3)
```

see where the station fall in relation to the GOA bottom trawl trawlable/untrawlable survey grid 

```{r}
# goa_stations_2025 <- st_read(here("data", "goa_stations_2025.gpkg"))
# 
# ggplot(goa_stations_2025) +
#   geom_sf(aes(fill = TRAWLABLE)) + 
#   scale_fill_manual(values = c("N" = "red", "Y" = "blue", "UNK" = "white")) +
#   theme_minimal()
```


add goa coral 2022 stations to this map 
```{r}
# # Create dissolved NMFS area boundaries
# nmfs_areas <- goa_stations_2025 %>%
#   group_by(NMFS_AREA) %>%
#   summarize(geometry = st_union(geom), .groups = "drop")
# 
# ggplot(goa_stations_2025) +
#   geom_sf(aes(fill = TRAWLABLE)) + 
#   scale_fill_manual(values = c("N" = "red", "Y" = "blue", "UNK" = "white")) +
#   # NMFS area outlines
#   geom_sf(
#     data = nmfs_areas,
#     fill = NA,
#     color = "black",
#     linewidth = 0.6) +
#   theme_minimal() + 
#   geom_sf(data = stations_sf_3338, size = 1, color = "green")
```

extract what the gap T/UT map says for each sampling station
also in include what strata
```{r}
# #check crs
# crs(goa_stations_2025)
# crs(stations_sf_3338)
# 
# # Spatial join: add TRAWLABLE values to points
# stations_joined <- st_join(
#   stations_sf_3338,
#   goa_stations_2025 %>% select(TRAWLABLE, NMFS_AREA),
#   join = st_within   # use st_intersects if needed
# )
# 
# stations_joined %>%
#   as.data.frame() %>%
#   count(NMFS_AREA, TRAWLABLE) %>%
#   tidyr::pivot_wider(
#     names_from = TRAWLABLE,
#     values_from = n,
#     values_fill = 0
#   )
```

okay, since there are lots of "unknown" cell let me try using the trawlable/untrawlable map from: https://www.sciencedirect.com/science/article/pii/S0165783623001789

```{r}
map <- read.csv(here("data", "grid_table_w_data_4.csv"))

map_sf <- st_as_sf(map, coords = c("long", "lat"), crs = 4326)

ggplot(map_sf) + 
  geom_sf(aes(color = score), size = 0.5) +
  #scale_color_viridis_c() +  
  scale_color_gradient2(low = "#BC4700", mid = "#D0D0D0", high = "#00467F", midpoint = 0, breaks = -4:5, labels = -4:5) + 
  theme_minimal() +
  labs(color = "Score")

ggplot(map_sf) + 
  geom_sf(aes(color = score2), size = 0.5) +
  scale_color_manual(values = c("#00467F", "#D0D0D0", "#BC4700")) + 
  theme_minimal() +
  labs(color = "Category")
```


```{r}
# Ensure same CRS as sampled stations
map_sf <- st_transform(map_sf, st_crs(stations_sf_3338))

ggplot(map_sf) + 
  geom_sf(aes(color = score), size = 0.5) +
  #scale_color_viridis_c() +  
  scale_color_gradient2(low = "#BC4700", mid = "#D0D0D0", high = "#00467F", midpoint = 0, breaks = -4:5, labels = -4:5) + 
  theme_minimal() +
  labs(color = "Score") + 
  geom_sf(data = stations_sf_3338, size = 1, color = "green")
```

extract nearest score
```{r}
# index of nearest map feature for each station
nearest_idx <- st_nearest_feature(stations_sf_3338, map_sf)

# extract nearest score
stations_sf_3338$score_nearest <- map_sf$score[nearest_idx]
```

extract mean of nearest neighbors
```{r}
library(nngeo)

k <- 5

# nearest neighbor indices
nn <- st_nn(
  stations_sf_3338,
  map_sf,
  k = k,
  returnDist = FALSE
)

# mean score across neighbors
stations_sf_3338$score_mean_k <- sapply(
  nn,
  function(idx) mean(map_sf$score[idx], na.rm = TRUE)
)
```

